Researchers preferentially collaborate with same-gendered colleagues across

### 2018-06-13 manuscript/Manuscript_Claire.Rmd
Small tweaks
Version for bioarXiv

---
title: "Researchers preferentially collaborate with same-gendered colleagues across the life sciences"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is stronger in fields where
  women are in the minority. Interestingly, journals with a high impact factor for their 
  discipline tended to have comparatively low homophily, as predicted if mixed-gender teams 
  produce better research. We discuss potential causes of gender homophily in science. 
  \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; studies listed in Table S1 of @Holman_2018]. For example, studies have concluded that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] or travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better outputs on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, multiple studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, all such studies imply that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing is often termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we name the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Whalund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if opposite-gendered colleagues are preferred. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers expected under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals (if anything, this small dataset suggests gender heterophily). The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Fig1.pdf)

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender. We also confirmed that the majority of papers had multiple authors, in most journals (S2 Data) and disciplines (S3 Data, S1 Fig).

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/Fig2.pdf)

@@

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, S4 Data). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, we cannot reject the null hypothesis that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

@@

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values than papers with more than two authors, on average (\autoref{author_number}; model output in Online Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers.

@@

## Relationship between journal impact factor and gender homophily
We observed a noisy but statistically significant linear relationship between standardised journal impact factor and $\alpha'$, such that journals with a high impact factor for their discipline had weaker gender homophily than did journals with a low impact factor for their discipline (\autoref{impact_factor}; linear regression: $R^2$ = `r R2`, $t_{1415}$ = -8.0, p < 0.0001).

@@

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (S4-S5 Fig). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); on average, $\alpha'$ was only inflated in the main (non-country specific) analysis by _c_. 0.002 (S6 Fig). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

@@

Despite this overlap, \autoref{simulation} suggests that our main conclusions (and those of other studies of gender homophily) are probably robust to this career stage issue. We only expect strongly positive $\alpha$ when A) the gender ratio is highly skewed across career stages (e.g. a 5-fold difference), and B) collaborations between early and established researchers are very rare (e.g. <10% of the total). Both of these conditions are untrue for most fields: the gender gap across careers stages is generally less pronounced [e.g. @Shaw_2012; @Holman_2018], and it is very common for early-career researchers to co-publish with an established mentor [@macaluso_2016]. However, one can get $\alpha > 0$ for realistic combinations of parameters, e.g. a moderate shortage of women in senior positions coupled with a moderate excess of within-career stage collaboration, suggesting this effect might contribute to some of the observed homophily.

@@

# Discussion
We found evidence that researchers preferentially publish with same-gendered coauthors, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite large: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, our findings are even more striking: for example, if men have 20% female coauthors and women have 30% (i.e. $\alpha' = 0.1$ in a field with a male-biased gender ratio), then women publish with women 50% more often than men do.

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconcious selection of same-gender collaborators by men, women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals; restricting the data by country yielded similar estimates of $\alpha'$; and we showed that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers do preferentially select same-gendered collaborators, although the effect size of this preference is difficult to measure.

We hypothesised that disciplines with a strongly skewed gender ratio might show the strongest gender homophily, e.g. because being in the minority might increase motivation to seek out same-gendered colleagues. Contrary to this hypothesis, we found no evidence that gender homophily is restricted to particular disciplines: $\alpha'$ was similarly high across the board (\autoref{alpha_histograms}). Moreover, $\alpha'$ was lower for journals with a male-biased author gender ratio relative to those with an even gender ratio. Interestingly, $\alpha'$ was highest in journals with a female-biased author gender ratio. This may suggest that men are more likely to preferentially seek out male collaborators in fields where men are a minority, relative to the homophily displayed by women in fields where women are a minority. However, this latter result is only tentatively supported since there are few journals in which most authors are women (\autoref{alpha_gender_ratio}).

@@

Regarding our finding of weaker homophily among 2-author papers, we suspect that many 2-author teams comprise a student/postdoc and a senior staff member, making these teams especially likely to be mixed-gender, due to the elevated gender gap among senior researchers [@Shaw_2012; @Holman_2018]. Assuming this interpretation is correct, this result suggests that our overall $\alpha'$ may underestimate the strength of peoples' preferences for same-gendered collaborators; essentially, women seeking a senior collaborator could be constrained to work mostly with men, meaning that people's ideal and realised gender preferences would be mismatched. On a related note, Ghiasi et al. [-@Ghiasi_2015] argue that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women; for reference, their Figure 7 suggests that coauthorships between women are roughly 30% more frequent than expected under random assortment. By contrast, we feel that it is not helpful to recommend that women collaborate primarily with other women, e.g. because this constrains women's options and may be counter-productive (particularly in fields like engineering, where 90% of professors are men [@Shaw_2012]). Instead, we suggest that researchers of both genders can help to close the gender gap in science. In the context of collaboration, one way to do this is to undertake self-examination to ensure that one is not inadvertently overlooking or excluding female potential students and colleagues. One should also take care to treat male and female collaborators equally, e.g. in terms of training and mentoring, allocation of work, and how one frames or promotes the collaboration (e.g. in conference presentations or on a website); evidence suggests that unconscious bias causes people to undervalue women's research achievements [@Moss_2012], and to assign menial or under-valued tasks to women and more prestigious and employable tasks (e.g. writing and project leadership) to men [@macaluso_2016].

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.

>>>>

---
title: Researchers preferentially collaborate with same-gendered colleagues across
  the life sciences
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: 
    keep_tex: true
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers preferentially co-publish with colleagues of the same
  gender, and show that this 'gender homophily' is slightly stronger today than it
  was 10 years ago. Contrary to our expectations, we found no evidence that homophily
  is driven mostly by senior academics, and no evidence that homophily is stronger
  in fields where women are in the minority. Interestingly, journals with a high impact
  factor for their discipline tended to have comparatively low homophily, as predicted
  if mixed-gender teams produce better research. We discuss some potential causes of gender
  homophily in science.
  \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Research collaboration, Text mining, Women in STEM.
---

@@

Writing papers, networking, and collaboration are all instrumental to research productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], and dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; studies listed in Table S1 of @Holman_2018]. For example, studies have concluded that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] or travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better outputs on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, multiple studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, most or all such studies imply that men co-publish with men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has often been termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we name the Wahlund effect (\autoref{wahlund_plot}), by analogy with the conceptually similar Whalund effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively difficult to infer gender-based preferences simply by counting the number of same- and mixed-gender coauthorships. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues even if no gender preferences exist, or if the true preference is for opposite-gendered colleagues ('gender heterophily'). For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers usually collaborate within their own discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers expected under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals (if anything, this small dataset suggests gender heterophily). The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. The example countries and disciplines were selected based on [@Holman_2018]. \label{wahlund_plot}](../figures/Fig1.pdf)

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. Most journals had positive values of $\alpha'$ (77-92%, depending on time period and author type; S1 Data), and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender. We also confirmed that the majority of papers had multiple authors, in most journals (S2 Data) and disciplines (S3 Data, S1 Fig).

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (corrected p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/Fig2.pdf)

@@

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had positive mean $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, S4 Data). However, there was no formal evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, the processes responsible for producing positive $\alpha'$ values appear to be similarly strong in all the disciplines we examined.

@@

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values relative to papers with more than two authors, on average (\autoref{author_number}; statistical results in Online Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers.

@@

## Relationship between journal impact factor and gender homophily
We observed a noisy but statistically significant linear relationship between standardised journal impact factor and $\alpha'$, such that journals with a high impact factor for their discipline had weaker gender homophily than did journals with a low impact factor for their discipline (\autoref{impact_factor}; linear regression: $R^2$ = `r R2`, $t_{1415}$ = -8.0, p < 0.0001). The slope of the regression was $-0.012{\pm}0.0015$, indicating that increasing the discipline-standardised impact factor by one standard deviation is associated with a reduction in $\alpha'$ of 0.012.

@@

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant heterophily (S4-S5 Fig). Additionally, the values of $\alpha'$ calculated for each journal-country combination were only very slightly lower than the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries, as was done to make \autoref{alpha_histograms}): on average, the difference in $\alpha'$ was only 0.002 (S6 Fig). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

@@

Despite this overlap, \autoref{simulation} suggests that our main conclusions (and those of other studies of gender homophily) are probably robust to this career stage issue. We only expect strongly positive $\alpha$ when A) the gender ratio is highly skewed across career stages (e.g. a 5-fold difference), and B) collaborations between early and established researchers are very rare (e.g. <10% of the total). Both of these conditions are untrue for most fields: the gender gap across careers stages is generally less pronounced [e.g. @Shaw_2012; @Holman_2018], and it is very common for early-career researchers to co-publish with an established mentor [@macaluso_2016]. However, one can get $\alpha > 0$ for realistic combinations of parameters, e.g. a moderate shortage of women in senior positions coupled with a moderate excess of within-career stage collaboration, suggesting this effect might contribute to some of the observed homophily (in this and previous studies).

@@

# Discussion
We found evidence that researchers preferentially publish with same-gendered coauthors, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite large: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, our findings are even more striking: for example, if men have 20% female coauthors and women have 30% (i.e. $\alpha' = 0.1$ in a field with a typical gender ratio [@Holman_2018]), then women publish with women 50% more often than men do.

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconcious selection of same-gender collaborators by men, women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals; restricting the data by country yielded similar estimates of $\alpha'$; and we showed that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers do preferentially select same-gendered collaborators, although the frequency and strength of this preference is difficult to ascertain.

We hypothesised that disciplines with a strongly skewed gender ratio might show the strongest gender homophily, e.g. because being in the minority might increase motivation to seek out same-gendered colleagues. Contrary to this hypothesis, we found no evidence that gender homophily is restricted to particular disciplines: $\alpha'$ was similarly high across the board (\autoref{alpha_histograms}). Interestingly, gender homophily was weakest for journals with a male-biased author gender ratio, and strongest in journals with a female-biased author gender ratio. This may suggest that men are more likely to preferentially seek out male collaborators in fields where men are a minority, relative to the homophily displayed by women in fields where women are a minority. However, this latter result is only tentatively supported since our sample contains few journals in which most authors are women (\autoref{alpha_gender_ratio}).

@@

Regarding our finding of weaker homophily among 2-author papers, we suspect that many 2-author teams comprise a student/postdoc and a senior staff member, making these teams especially likely to be mixed-gender, due to the elevated gender gap among senior researchers [@Shaw_2012; @Holman_2018]. Assuming this interpretation is correct, this result suggests that our reported $\alpha'$ values may underestimate the strength of peoples' preferences for same-gendered collaborators; essentially, women seeking a senior collaborator could be constrained to work mostly with men, meaning that people's ideal and realised gender preferences would be mismatched. On a related note, Ghiasi et al. [-@Ghiasi_2015] argue that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women (their Figure 7 suggests that coauthorships between women are 30% more frequent than expected under random assortment). By contrast, we feel that it is not helpful to recommend that women collaborate primarily with other women, e.g. because this constrains women's options and may be counter-productive (particularly in fields like engineering, where 90% of professors are men [@Shaw_2012]). Instead, we suggest that researchers of both genders can help to close the gender gap in STEMM. In the context of collaboration, one way to do this is to undertake self-examination to ensure that one is not inadvertently overlooking or excluding female potential students and colleagues. One should also take care to treat male and female collaborators equally, e.g. in terms of training and mentoring, allocation of work, and how one frames or promotes the collaboration (e.g. in conference presentations or on a website); evidence suggests that unconscious bias causes people to undervalue women's research achievements [@Moss_2012], and possibly to assign menial or under-valued tasks to women and more prestigious tasks to men [@macaluso_2016].

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to gender differences in confidence, or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.
